### 2023-05-10 writing/manuscript.Rmd
fixing one small discrepancy to final published version

The lexical use of pitch in tonal languages is distinct from how pitch is otherwise used in speech. For example, many languages use pitch to convey affect [@Cowen2019]; to cue non-lexical meaning [e.g., helping to differentiate between questions and statements\; @Patel2008; @Tong2005]; to emphasize information [@Breen2010]; to cue sentence structure with metrical stress patterns [@Wagner2019], supporting comprehension [@Hilton2021b]; and/or as a cue to speech categories, as in infant- or child-directed speech [@Hilton2022a]. While these many uses of pitch are typical of speech in both tonal and non-tonal languages [e.g., many Indo-European, South Asian, or Australian languages\; @Maddieson2013], in non-tonal languages, pitch is never used lexically to denote word meanings. Lastly, *pitch-accented* languages form an intermediate category with limited or mixed use of lexical pitch [such as Croatian\; @vanderHulst2011]; note, however, that whether pitch-accented languages form a coherent standalone category or whether they are better considered on a spectrum between tonal and non-tonal languages, with some mixed cases, is a matter of debate [e.g., @Gussenhoven2004; @Hyman2006; @Hyman2009].^[Indeed, a category of languages that groups evidently disparate languages together, such as Japanese and Swedish, may not be defensible. As pitch-accented languages are not our primary focus, here we treat them as a separate group from tonal and non-tonal languages, but also conduct some analyses at the language level rather than the language-type level.]

@@

We aggregated summary information from `r study_total_num` prior studies of music perception and tonal language experience (see SI Text 1.1 for the search criteria) and studied them with random-effects meta-analysis models. Because there is no consensus in the literature about what specific music processing ability to expect a tonal-language advantage on, we grouped the prior studies into three rough categories: melody, fine-grained pitch, and rhythm. We built separate meta-analytic models for each group.

@@

In the rest of this paper, we report a study of the ability to discriminate melodies, detect mistuned singing, and detect misaligned beats, in `r f(users$final)` people across the globe. Participants self-reported their native language, location, demographics, and degree of musical training, enabling language-wise and language-type-wise analyses of each of these musical abilities and their relations to linguistic and musical experience.

@@

Generalization from particular speakers of specific languages to entire groups of languages, with high statistical power, and with controls for potential sources of systematic bias, requires large samples of participants representing a variety of languages and countries [@Blasi2022; @Yarkoni2022]. We report such a test here, using methods of gamification and citizen science [@Hilton2022; @Huber2020; @Hartshorne2019; @Long2023].

@@

Contrasting with the prior meta-analytic effects, however, there was no clear advantage for tonal-language speakers on the vocal mistuning task ($\beta$ = `r main_analysis_estimates$mpt$ToneTonal$beta`, $t$ = `r main_analysis_estimates$mpt$ToneTonal$t`, $p$ = `r main_analysis_estimates$mpt$ToneTonal$p`; Table 2), despite the fact that the task *did* reliably demonstrate clear performance differences between those with and without music training ($\beta$ = `r main_analysis_estimates$mpt$musicLessonsYes$beta`, $t$ = `r main_analysis_estimates$mpt$musicLessonsYes$t`, $p$ `r main_analysis_estimates$mpt$musicLessonsYes$p`). In other words, while some experiences (e.g., musical training) do shape fine-grained pitch perception in the context of vocal mistuning, there seems to be minimal or no effect of tonal-language experience on this ability. This result contradicts prior studies of fine-grained pitch discrimination that reported tonal-language advantages in mainly Mandarin and Cantonese speakers.

Lastly, contrary the meta-analytic results, which found no effect of tonal language experience on rhythm perception, native speakers of tonal languages performed *worse* in beat perception relative to the non-tonal and pitch-accented groups ($\beta$ = `r main_analysis_estimates$cabat$ToneTonal$beta`, $t$ = `r main_analysis_estimates$cabat$ToneTonal$t`, $p$ `r main_analysis_estimates$cabat$ToneTonal$p`). Like the melodic discrimination effect, the size of this effect was large, approaching the size of effects of having taken music lessons (see Figure 2).

@@

While the scope of our data collection allowed for analysis of music processing abilities in thousands of native speakers of six pitch-accented languages, the findings concerning these speakers were murky and failed to replicate across different analysis approaches. Within-language group variability was also high (see Figure 4) for pitch-accented languages, suggesting no common group advantage/disadvantage across speakers of pitch-accented languages. This, of course, is complicated by the inherently fuzzier nature of classification of pitch-accented languages, relative to tonal languages (see Footnote 1)[@Gussenhoven2004; @Hyman2006; @Hyman2009]. Further work that codifies that nature of both linguistic and musical pitch use across this set of generally understudied languages may provide more clarity. We encourage interested readers to re-analyze the open-access data reported here, using alternative classifications of languages into the "pitch-accented" category.

>>>>

The lexical use of pitch in tonal languages is distinct from how pitch is otherwise used in speech. For example, many languages use pitch to convey affect [@Cowen2019]; to cue non-lexical meaning [e.g., helping to differentiate between questions and statements\; @Patel2008; @Tong2005]; to emphasize information [@Breen2010]; to cue sentence structure with metrical stress patterns [@Wagner2019], supporting comprehension [@Hilton2021b]; and/or as a cue to speech categories, as in infant- or child-directed speech [@Hilton2022a]. While these many uses of pitch are typical of speech in both tonal and non-tonal languages [e.g., many Indo-European, South Asian, or Australian languages\; @Maddieson2013], in non-tonal languages, pitch is never used lexically to denote word meanings. Lastly, *pitch-accented* languages form an intermediate category with limited or mixed use of lexical pitch [such as Croatian\; @vanderHulst2011]; note, however, that whether pitch-accented languages form a coherent standalone category or whether they are better considered on a spectrum between tonal and non-tonal languages, with some mixed cases, is a matter of debate [e.g., @Gussenhoven2004; @Hyman2006; @Hyman2009]. As pitch-accented languages are not our primary focus, here we treat them as a separate group from tonal and non-tonal languages, but also conduct some analyses at the language level rather than the language-type level.

@@

We aggregated summary information from 20 prior studies of music perception and tonal language experience (see SI Text 1.1 for the search criteria) and studied them with random-effects meta-analysis models. Because there is no consensus in the literature about what specific music processing ability to expect a tonal-language advantage on, we grouped the prior studies into three rough categories: melody, fine-grained pitch, and rhythm. We built separate meta-analytic models for each group.

@@

Thus, the meta-analysis demonstrates evidence for a potential effect of tonal language experience on melodic discrimination ability and finer-grained pitch discrimination ability, but a lack of linguistic diversity, low statistical power, and high likelihood of systematic bias in the samples studied warrant caution and call into question the reproducibility and generalizability of these findings [@Blasi2022; @Yarkoni2022]. These issues can be addressed by studying many native speakers of many languages, from different countries, with and without music training experience, all of whom complete the same assessments of music processing ability.

@@

We report such a test here, using methods of gamification and citizen science [@Hilton2022; @Huber2020; @Hartshorne2019; @Long2023] to recruit `r f(users$final)` people from 203 countries across the globe, including `r f(users$speakers$Tonal)` native speakers of `r f(users$langs$Tonal)` tonal languages, `r f(users$speakers$Pitch_accented)` native speakers of `r f(users$langs$Pitch_accented)` pitch-accented languages, and `r f(users$speakers$Non_tonal)` native speakers of `r f(users$langs$Non_tonal)` non-tonal languages (see Figure 1).

@@

Also consistent with the meta-analysis, there was no clear advantage for tonal language speakers on the vocal mistuning task ($\beta$ = `r main_analysis_estimates$mpt$ToneTonal$beta`, $t$ = `r main_analysis_estimates$mpt$ToneTonal$t`, $p$ = `r main_analysis_estimates$mpt$ToneTonal$p`; Table 1), despite the fact that the task did reliably demonstrate clear performance differences between those with and without music training ($\beta$ = `r main_analysis_estimates$mpt$musicLessonsYes$beta`, $t$ = `r main_analysis_estimates$mpt$musicLessonsYes$t`, $p$ `r main_analysis_estimates$mpt$musicLessonsYes$p`). In other words, while some experiences (e.g., musical training) do shape fine-grained pitch perception in the context of vocal mistuning, there seems to be minimal or no effect of tonal language experience on this ability.

Last, contrary to the meta-analytic results, which found no effect of tonal language experience on rhythm perception, native speakers of tonal languages performed worse in beat perception relative to the non-tonal and pitch-accented groups ($\beta$ = `r main_analysis_estimates$cabat$ToneTonal$beta`, $t$ = `r main_analysis_estimates$cabat$ToneTonal$t`, $p$ `r main_analysis_estimates$cabat$ToneTonal$p`). Like the melodic discrimination effect, the size of this effect was large, approaching the size of effects of having taken music lessons (see Figure 2).

@@

While the scope of our data collection allowed for analysis of music processing abilities in thousands of native speakers of six pitch-accented languages, the findings concerning these speakers were murky and failed to replicate across different analysis approaches. Within-language group variability was also high (see Figure 4) for pitch-accented languages, suggesting no common group advantage/disadvantage across speakers of pitch-accented languages. This, of course, is complicated by the inherently fuzzier nature of classification of pitch-accented languages, relative to tonal languages [@Gussenhoven2004; @Hyman2006; @Hyman2009]. Further work that codifies that nature of both linguistic and musical pitch use across this set of generally understudied languages may provide more clarity. We encourage interested readers to re-analyze the open-access data reported here, using alternative classifications of languages into the "pitch-accented" category.
